Detecting confounder and center bias in machine learning models

### 2021-10-26 manuscript.tex
Updates from Overleaf

% tab:conditional-independence-cases
\renewcommand{\arraystretch}{1.2}
\begin{table}[]
\centering
\begin{tabular}{l|rp{60mm}|c|>{\centering\arraybackslash}m{30mm}}
 &  & H0  & assumption needed for: & no assumptions about the distribution of: \\
\hline
\textbf{1.} & $\y \independent \yhat | \c$ \quad  & full confounder test: model exclusively driven by confounder & $Q(\cdot|\c)$ & $\yhat$ or $\y$ \\
%\hline
\textbf{2.} & $\y \independent \c | \yhat$ \quad & model captures all variance in confounder (not of interest) & $Q(\cdot|\yhat)$ & $\y$ or $\c$ \\
%\hline
\textbf{3.} & $\yhat \independent \c | \y$  \quad &  partial confounder  test: model does not capture more confound than what is present in the target & $Q(\cdot|\y)$ & $\yhat$ or $\c$ \\
\end{tabular}
\caption{\label{tab:conditional-independence-cases} Possibilities when testing conditional independence in potentially biased predictive models. \\The table lists the three possible null hypotheses (H0), and the variables where assumption about the joint/conditional distributions is required/not required.   ($y$: prediction target, $\yhat$: predictions, $\c$: confounder variable) }
\end{table}

@@

The partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\c$,

@@

where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that

@@

is a permutation of the original vector $\c = (c_1, \dots, c_n)$, with its elements reordered according to the permutation $\quad \boldsymbol{\pi} \in S_n$ where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$.

@@

While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times. The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data. The acceptance odds ratio of swapping indices $i$ and $j$ is:

@@

Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $c$ and the confounder and the target signals in $\yhat$).
Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$

@@

$$ \c \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yc} \quad g(\y) $$

where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} defiend as:

@@

Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced) or e.g. a sigmoid-shaped function, in pour case:

@@

$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yyhat} \quad g(\y)  + w_{c\hat{y}} \quad g(\y)$$

@@

To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing, its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq. \ref{eq:linreg}, as described by \citep{berrett2020conditional}) in the following parameters:
\begin{itemize}
    \item sample size: $n = 1000$
    \item $w_{c\hat{y}} = 0$ (H0 simulations only)
    \item all combinations of:
    \begin{itemize}
        \item $w_{yc} \in \{0.5, 1, 2, 3\}$ and
        \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
    \end{itemize}
    \item non-normality ($f_{\delta = 0.1, \epsilon = 2)}$ and non-linearity (sigmoid $g$)
\end{itemize}

>>>>

% tab:conditional-independence-cases
\renewcommand{\arraystretch}{1.2}
\begin{table}[]
\centering
\begin{tabular}{l|rp{60mm}|c|>{\centering\arraybackslash}m{30mm}}
 &  & H0  & assumption needed for: & no assumptions about the distribution of: \\
\hline
\textbf{1.} & $\yhat \independent \y | \c$ \quad  & full confounder test: model exclusively driven by the confounder & $Q(\y|\c)$ & $(\yhat,\y), (\yhat, \c)$ \\
%\hline
\textbf{2.} & $\y \independent \c | \yhat$ \quad & model captures all variance in the confounder (not of interest) & $Q(\c|\yhat)$ & $(\y, \c), (\y, \yhat)$ \\
%\hline
\textbf{3.} & $\yhat \independent \c | \y$  \quad &  partial confounder  test: model not directly driven by the confoudner & $Q(\c|\y)$ & $(\yhat, \c), (\yhat, \y)$ \\
\end{tabular}
\caption{\label{tab:conditional-independence-cases} Possibilities when testing conditional independence in potentially biased predictive models. \\The table lists the three possible null hypotheses (H0), and the variables where assumption about the joint/conditional distributions is required/not required.   ($\y$: prediction target, $\yhat$: predictions, $\c$: confounder variable) }
\end{table}

@@

The partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based 'copy' of $\c$,

@@

where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y_i$ and $j=1,\dots, m$ indexes the 'copy' of $\c$ so that

@@

is a permutation of the original vector $\c = (c_1, \dots, c_n)$, with its elements reordered according to the permutation $ \boldsymbol{\pi} \in S_n$ where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$.

@@

The required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random. However, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times. The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data. The acceptance odds ratio of swapping indices $i$ and $j$ is:

@@

Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $c$ and the confounder and the target signals in $\yhat$).
Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$.

@@

$$ \c \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yc} \, g(\y) $$

where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} definend as:

@@

Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced) or e.g. a sigmoid-shaped function, in our case:

@@

$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yyhat} \: g(\y) \, + \, w_{c\hat{y}} \: g(\y)$$

@@

To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing, its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq. \ref{eq:linreg}, as described by \cite{berrett2020conditional}) with the following simulation parameters:
\begin{itemize}
    \item sample size: $n = 1000$
    \item $w_{c\hat{y}} = 0$ (H0 simulations only)
    \item all combinations of:
    \begin{itemize}
        \item $w_{yc} \in \{0.5, 1, 2, 3\}$ and
        \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
    \end{itemize}
    \item non-normality ($f_{\delta = 0.1, \epsilon = 2)}$ and non-linearity (sigmoid $g$)
\end{itemize}
